---
title: "Final report: COPD rates compared to average air quality throughout the year"
author: "Jacob Peterson"
date: "10/8/2025"
output: github_document
---

- [ABSTRACT](#abstract)
- [INTRO](#intro)
- [BACKGROUND](#background)
- [STUDY QUESTION and HYPOTHESIS](#study-question-and-hypothesis)
  - [Questions](#questions)
  - [Hypothesis](#hypothesis)
  - [Prediction](#prediction)
- [METHODS](#methods)
  - [Fill in 1st analysis
    e.g. barplots](#fill-in-1st-analysis-eg-barplots)
  - [Fill in 2nd analysis/plot e.g. generalized linear
    model](#fill-in-2nd-analysisplot-eg-generalized-linear-model)
- [DISCUSSION](#discussion)
  - [Interpretation of 1st analysis
    (e.g. barplots)](#interpretation-of-1st-analysis-eg-barplots)
  - [Interpretation of 2nd analysis (e.g. generalized linear
    model)](#interpretation-of-2nd-analysis-eg-generalized-linear-model)
- [CONCLUSION](#conclusion)
- [REFERENCES](#references)

# ABSTRACT


# Intro



# BACKGROUND

Air Quality Index (AQI) is a standardized system used to measure the quality of the air around us. The AQI is calculated using the four major air pollutants, as determined by the Clean Air Act, which are ground-level ozone, particle pollution, carbon monoxide, and sulfur dioxide. AQI is measured on a scale of 0-500, with 0 being the cleanest air and 500 being the most hazardous. This categorizes 0 as the highest air quality and 500 as the lowest air quality. AQI is a useful tool that has become integrated into our weather apps, making it accessible to the majority of the population.
One of the primary causes of death in urban areas is chronic obstructive pulmonary disease (COPD). COPD can be caused by many factors including tobacco use, occupational factors, infection, and air pollution. This disease progresses slowly and worsens with prolonged exposure. The damage is usually not reversible and can lead to pulmonary failure. There are many factors that play into the prevalence of COPD like region, age, and sex.  


# STUDY QUESTION and HYPOTHESIS

With this information in mind, we have posed the following question: “Do areas of low air quality have higher rates of Chronic Obstructive Pulmonary Disease (COPD)?” Although there are many factors that influence COPD, if there is a negative correlation between air quality and rates of COPD there are many known preventative measures that can be taken to prevent COPD and improve lung health. These findings can influence health practices and help those susceptible to COPD avoid risk factors, such as low air quality. Although there are many factors that can play into rates of COPD, we hypothesize that this negative correlation will exist and that areas with lower air quality will have higher rates of Chronic Obstructive Pulmonary Disease.

## Questions


## Hypothesis

## Prediction



# METHODS



```{r setup, include=FALSE}
#packages
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(viridis)
library(maps)
```
```{r include = FALSE}

data_text <- "
StatePercentCountPercentCountPercentCount
Alabama 8.6 161,900 10.1 207,500 9.4 369,400
Alaska 5.3 15,300 6.1 16,100 5.7 31,400
Arizona 5.0 137,800 6.4 180,700 5.7 318,500
Arkansas 8.6 97,600 10.5 125,600 9.6 223,200
California 4.5 680,600 4.7 733,100 4.6 1,413,800
Colorado 4.3 98,900 5.9 135,100 5.1 233,900
Connecticut 4.7 65,400 5.9 87,200 5.3 152,500
Delaware 5.9 22,400 6.6 27,300 6.2 49,700
District of Columbia 4.1 10,600 5.1 15,100 4.6 25,700
Florida (2020) 6.8 574,600 8.2 745,500 7.5 1,320,100
Georgia 5.3 211,000 7.9 341,100 6.6 552,100
Hawaii 3.4 19,300 3.6 20,800 3.5 40,200
Idaho 5.0 35,400 6.4 45,800 5.7 81,200
Illinois 5.4 260,600 5.6 287,300 5.5 547,900
Indiana 7.4 188,300 9.5 256,200 8.5 444,400
Iowa 6.0 72,500 7.1 89,000 6.6 161,600
Kansas 5.9 64,400 7.0 79,300 6.4 143,700
Kentucky 9.8 166,800 12.0 214,900 10.9 381,700
Louisiana 8.5 144,500 8.9 163,700 8.7 308,200
Maine 8.1 43,600 9.8 56,200 9.0 99,900
Maryland 3.7 85,000 5.9 149,000 4.8 234,000
Massachusetts 5.0 136,300 6.7 196,200 5.9 332,500
Michigan 6.8 261,800 8.6 348,500 7.7 610,200
Minnesota 5.0 108,600 4.4 98,700 4.7 207,400
Mississippi 7.6 82,500 10.4 122,400 9.1 204,900
Missouri 6.6 153,500 10.4 255,600 8.5 409,000
Montana 5.2 22,800 6.6 28,400 5.9 51,200
Nebraska 5.2 38,300 6.1 45,600 5.6 83,900
Nevada 5.7 69,200 8.0 98,600 6.8 167,800
New Hampshire 4.7 165,200 6.3 235,000 5.5 400,100
New Jersey 6.3 35,100 8.1 46,400 7.2 81,400
New Mexico 4.5 36,100 6.7 55,900 5.6 91,900
New York 5.0 379,600 5.9 490,200 5.5 869,900
North Carolina 6.1 241,000 9.0 386,500 7.6 627,500
North Dakota 4.4 13,500 5.3 15,600 4.8 29,100
Ohio 7.6 340,300 9.5 452,900 8.6 793,200
Oklahoma 6.6 98,500 8.9 136,800 7.8 235,300
Oregon 5.1 85,200 7.0 120,300 6.1 205,500
Pennsylvania 6.9 344,000 6.8 361,300 6.8 705,300
Rhode Island 5.8 24,600 5.7 26,000 5.7 50,600
South Carolina 6.7 130,600 8.3 177,200 7.5 307,800
South Dakota 5.7 19,300 6.2 20,800 6.0 40,100
Tennessee 9.7 254,600 11.1 311,100 10.4 565,700
Texas 5.2 570,900 6.8 771,100 6.0 1,342,000
Utah 3.9 46,500 4.7 56,300 4.3 102,800
Vermont 5.8 15,000 7.3 19,700 6.6 34,700
Virginia 5.4 179,600 7.3 254,500 6.4 434,100
Washington 4.7 143,400 5.6 172,300 5.1 315,700
West Virginia 11.8 82,300 14.4 103,900 13.1 186,200
Wisconsin 5.5 126,400 5.7 132,800 5.6 259,200
Wyoming 6.2 14,100 6.9 15,100 6.5 29,200
United States 5.5 6,800,400 6.8 8,875,100 6.2 15,587,000
"
```

```{r echo = FALSE}
lines <- str_split(data_text, "\n")[[1]]
lines <- trimws(lines)
lines <- lines[lines != ""]
lines <- lines[-1]

# 4. Define the pattern
pattern <- "^([A-Za-z .()0-9]+?)\\s+(\\d+\\.\\d)\\s+([\\d,]+)\\s+(\\d+\\.\\d)\\s+([\\d,]+)\\s+(\\d+\\.\\d)\\s+([\\d,]+)$"
rows <- str_match(lines, pattern)
rows <- rows[!is.na(rows[,1]), ]

# 7. Create the data frame
df <- as.data.frame(rows[,2:8])
colnames(df) <- c("State","Percent_Male","Count_Male","Percent_Female","Count_Female","Percent_Total","Count_Total")
df <- df %>%
  # Filter out rows i don't want to plot
  filter(State != "United States") %>%
  filter(State != "District of Columbia") %>% # <-- NEW FIX: Remove D.C.
  # Convert columns to the right type
  mutate(across(starts_with("Percent"), as.numeric),
         across(starts_with("Count"), ~ as.numeric(str_replace_all(.x, ",", ""))),
         
         State = str_replace(State, "\\s*\\(.*\\)", ""), # Removes (2020)
         State = str_trim(State))                       # Removes whitespace
```
```{r echo = FALSE}
# 1. map data
states_map <- map_data("state")

# 2. heat map data to join
heatmap_data <- df %>%
  select(State, Percent_Female, Percent_Male) %>%
  pivot_longer(cols = starts_with("Percent_"),
               names_to = "Group",
               values_to = "Percent") %>%
  mutate(Group = str_replace(Group, "Percent_", ""),
         region = tolower(State)) 
map_plot_data <- left_join(states_map, heatmap_data, by = "region", relationship = "many-to-many")
# NEW stuff this better work:
map_plot_data %>%
  filter(!is.na(Group)) %>% 
  ggplot(aes(x = long, y = lat, group = group, fill = Percent)) +
  
  geom_polygon(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Percent") +
  facet_wrap(~ Group) + 
  labs(title = "COPD Prevalence by U.S. State and Gender") +
  theme_void() + 
  coord_map() + 
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    strip.text = element_text(size = 14, face = "bold")
  
  )
```

```{r echo = FALSE}

states_map <- map_data("state") 
difference_data <- df %>%
  mutate(
    Percent_Difference = Percent_Female - Percent_Male,
    region = tolower(State) # <-- This creates the join key
  ) %>%
  select(region, Percent_Difference)
map_diff_data <- left_join(states_map, difference_data, by = "region")
ggplot(map_diff_data, aes(x = long, y = lat, group = group, fill = Percent_Difference)) +
  geom_polygon(color = "white", linewidth = 0.1) + # Draw the states
  # 5. Use the DIVERGING color scale
  scale_fill_gradient2(
    name = "Percent Difference\n(Female - Male)",
    low = "blue",      # States where males are higher (negative)
    mid = "white",     # States where rates are equal (zero)
    high = "red",      # States where females are higher (positive)
    midpoint = 0       # We center the scale at zero
  ) +
  
  labs(title = "Difference in COPD Prevalence (Female vs. Male)") +
  theme_void() + # Use a clean theme
  coord_map() +  # Use correct map projection
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```


# air quality index
```{r include = FALSE}
# more packages
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(viridis)
library(maps)
library(readr) # Added to read the CSV
```

```{r include=FALSE}
aq_df <- read_csv("air-quality-by-state-2025.csv")

print("--- Air Quality Data ---")

```

```{r echo = FALSE}
states_map <- map_data("state")

aq_data_to_plot <- aq_df %>%
  # Use the correct column names: `state` and `AirQuality_AirQualityIndexViaUSA_num_YearFree`
  # We also rename the long AQI column to `Overall_AQI` to make it easier to use
  select(State = state, Overall_AQI = `AirQuality_AirQualityIndexViaUSA_num_YearFree`) %>% 
  mutate(region = tolower(State)) # <-- This creates the join key

map_plot_data <- left_join(states_map, aq_data_to_plot, by = "region")


map_plot_data <- map_plot_data %>%
  filter(!is.na(Overall_AQI))

ggplot(map_plot_data, aes(x = long, y = lat, group = group, fill = Overall_AQI)) +
  geom_polygon(color = "white", linewidth = 0.1) + # Draw the states
  
  scale_fill_viridis_c(name = "Overall AQI") +
  
  labs(title = "Overall Air Quality Index (AQI) by U.S. State (2025)") +
  theme_void() + # Use a clean theme
  coord_map() +  # Use correct map projection
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

``` {r include = FALSE, eval = FALSE}
> ggplot(df, aes(x = sqrt(Percent_Total))) +
+     geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") +
+     labs(
+         title = "Normalized Distribution of COPD Prevalence",
+         x = "Square Root of Overall COPD Rate",
+         y = "Number of States"
+     ) +
+     theme_minimal() +
+     theme(
+         plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
+     )
> # We must use backticks for the column name
> aqi_column_name <- "AirQuality_AirQualityIndexViaUSA_num_YearFree"
> 
> # Use mutate() to create the new transformed column
> aq_df_normalized <- aq_df %>%
+     mutate(
+         # 1. Find the max value (k)
+         k = max(!!sym(aqi_column_name)) + 1,
+         
+         # 2. "Invert" (Reflect) and "SQRT"
+         # The formula is sqrt(k - x)
+         AQI_Normalized = sqrt(k - !!sym(aqi_column_name))
+     )
> 
> # Now, plot the new normalized column
> ggplot(aq_df_normalized, aes(x = AQI_Normalized)) +
+     geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black") + 
+     labs(
+         title = "Normalized Distribution of AQI (Reflected + SQRT)",
+         x = "Square Root of (Max AQI - AQI)",
+         y = "Number of States"
+     ) +
+     theme_minimal() +
+     theme(
+         plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
+     )




# DISCUSSION



# CONCLUSION


